Efficient Unconstraining Parameter Transforms for Hamiltonian Monte Carlo

### 2022-07-27 dummypaper.tex
figures

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{cite}
\usepackage[letterpaper, total={6.5in, 9in}]{geometry}
\usepackage{mathpazo}
\usepackage{sourcecodepro}

@@

\section{Changes of Variables}

@@

\section{Preliminaries}

[SA: TODO: eliminate or merge with following section]

@@

\section{Computing Jacobian determinants}

[SA: TODO: remove everything we do not use later, add strategy for making non-square Jacobian square]

@@

When the previously discussed strategies are insufficient to compute the Jacobian, we bypass the Jacobian computation entirely and use the exterior product $\wedge$.
For example, let $X \in \mathbb{R}^{N \times K}$.
The exterior product
\[
  \dv{X_{11} }\wedge \dv{X_{21}} \wedge \ldots \wedge \dv{X_{NK}} = \bigwedge_{j=1}^K \bigwedge_{i=1}^N \dv{X}_{ij}
\]
is another more explicit way of writing the integration measure $\dv{X_{11}} \dv{X_{21}} \dots \dv{X_{NK}}$ that has useful properties for computing Jacobian determinants.
The basic properties we will use are

\begin{itemize}
  \item $a\dv{x} \wedge b\dv{y} = (ab)(\dv{x} \wedge \dv{y})$ for $a,b \in \mathbb{R}$
  \item $(\dv{x} + \dv{y}) \wedge \dv{z} = \dv{x} \wedge \dv{z} + \dv{y} \wedge \dv{z}$
  \item $\dv{x} \wedge \dv{y} = - \dv{y} \wedge \dv{x}$
  \item $\dv{x} \wedge \dv{x} = 0$
\end{itemize}

The connection between the exterior product and the Jacobian determinant is given by the following expression.
Given $f: y \mapsto x$ for $x,y \in \mathbb{R}^N$,
\[\bigwedge_{i=1}^N \dv{x_i} = \det(J_f(y)) \bigwedge_{i=1}^N \dv{y_i}.\]

We can then find the Jacobian determinant by first writing all differential forms $\dv{x_i}$ or $\dv{y_i}$ and then computing the exterior product until the expression looks like the above.
For our purposes, we will ignore the sign of the Jacobian determinant.

A related useful property is that for $X,Y,B \in \mathbb{R}^{N \times K}$, if $\dv{X} = B \dv{Y}$, then
\[\bigwedge_{j=1}^K \bigwedge_{i=1}^N \dv{X_i} = \det(B)^K \bigwedge_{j=1}^K \bigwedge_{i=1}^N \dv{Y_i}\]

Following [REF, Edelman], we use the notation $(\cdot)^\wedge$ as a shorthand to represent the exterior product of the unique elements of $\cdot$.

For example,
\begin{itemize}
  \item Rectangular matrix $\{X \in \mathbb{R}^{N \times K}\}$: $(\dv{X})^\wedge = \bigwedge_{j=1}^K \bigwedge_{i=1}^N \dv{X_{ij}}$
  \item Lower-triangular rectangular matrix $\{X \in \mathbb{R}^{N \times K} | X_{ij} = 0\ \forall\ i < j\}$: $(\dv{X})^\wedge = \bigwedge_{i=1}^N \bigwedge_{j=1}^{\min(i, K)} \dv{X_{ij}}$
  \item Diagonal matrix $\{X \in \mathbb{R}^{N \times K} | X_{ij} = 0\ \forall\ i \ne j\}$: $(\dv{X})^\wedge = \bigwedge_{i=1}^{\min(N, K)} \dv{X_{ii}}$
  \item Symmetric matrix $\{X \in \mathbb{R}^{N \times N} | X_{ij} = X_{ji}\}$: $(\dv{X})^\wedge = \bigwedge_{i=1}^N \bigwedge_{j=1}^i \dv{X_{ij}}$
  \item Skew-symmetric matrix $\{X \in \mathbb{R}^{N \times N} | X_{ij} = -X_{ji}, X_{ii} = 0\}$: $(\dv{X})^\wedge = \bigwedge_{i=1}^N \bigwedge_{j=1}^{i-1} \dv{X_{ij}}$
\end{itemize}

Using this notation, for example, $(B \dv{X})^\wedge = \det(B)^K (\dv{X})^\wedge$

@@

\section{Semi-orthogonal matrices (Stiefel manifold)}

A matrix $X \in \mathbb{R}^{N \times K}$ for $K \le N$ is semi-orthogonal if and only if $X^\top X = I_K$.
The semi-orthogonal matrices form a space called the Stiefel manifold.
When $K=1$, $X$ is a point on the $(N-1)$-sphere $\mathbb{S}^{N-1} \subset \mathbb{R}^N$.
We will consider such unit vectors in the next section.
When $K=N$, $X$ is an element of the orthogonal group $O(N)$ and additionally satisfies $\det(X) = \pm 1$.
$O(N)$ is comprised of two disconnected submanifolds, one being the Special Orthogonal group $SO(N)$ with $\det(X) = +1$, and the other having $\det(X) = -1$.
$SO(2)$ and $SO(3)$ are the groups of rotation matrices of 2D and 3D vectors, respectively.

By choosing a set of rules such as Gram-Schmidt orthonormalization of the canonical basis in $\mathbb{R}^{N \times N}$, it is possible to analytically compute from $X$ an orthogonal complement $X^\perp \in \mathbb{R}^{N \times (N - K)}$, such that $\tilde{X}^\top \tilde{X} = I_N$, where $\tilde{X} = \begin{bmatrix}X & X^\perp \end{bmatrix} \in O(N)$.

Note that if we differentiate both sides of the constraint $X^\top X = I_K$, we get $\dv{X}^\top X + X^\top \dv{X} = 0$, which leads to $(X^\top \dv{X})^\top = - X^\top \dv{X}$.
Therefore, $X^\top \dv{X}$ is a $K \times K$ skew-symmetric matrix with $\frac{K(K-1)}{2}$ unique entries in the strict upper triangle.

Densities of distributions on the Stiefel manifold are usually written with respect to the invariant metric on the Stiefel manifold, which we will define separately for the two cases $K=N$ and $K < N$.

For $X \in O(N)$ with the $i$th column written $x_i$, we define the invariant measure as the exterior product of the unique elements of this skew-symmetric matrix:
\[
  (X^\top \dv{X})^\wedge \equiv \bigwedge_{j=1}^N \bigwedge_{i=j+1}^N (X^\top \dv{X})_{ij} = \bigwedge_{j=1}^N \bigwedge_{i=j+1}^N x_i^\top \dv{x}_j.
\]
This measure is called invariant because if $f(X) = A X B$ for $A,B \in O(N)$ and $\mathcal{S} \subset O(N)$ then $\int_{\mathcal{S}} (X^\top \dv{X}) = \int_{f(\mathcal{S})} (X^\top \dv{X})$.
It is also called the Haar measure on $O(N)$ \cite{muirhead2009aspects}..

To choose an invariant measure for $K < N$, we first choose an $X^\perp$ analytically computed from $X$.
Because $X^\perp$ is analytically computed, $\dv{X^\perp_{ij}} = 0$.
We can then construct the invariant measure for $K<N$ from the invariant measure for $O(N)$:

\[
  (X^\top \dv{X})^\wedge \equiv \bigwedge_{j=1}^N \bigwedge_{i=j+1}^N x_i^\top \dv{x}_j = \bigwedge_{j=1}^K \bigwedge_{i=j+1}^N x_i^\top \dv{x}_j,
\]

Though we needed to choose a $X^\perp$, the invariant measure does not depend on the choice of $X^\perp$.
Moreover, this measure is invariant to left-actions by $O(N)$ and right-actions by $O(K)$.
For more details, see \cite{muirhead2009aspects}.

When $K=1$, the invariant measure is equivalent to the usual Hausdorff measure on the sphere.

\subsection{Givens parameterization of Stiefel manifold}

\subsection{Cayley parameterization of Stiefel manifold}

\subsection{Polar parameterization of Stiefel manifold}

\subsection{Householder parameterization of Stiefel manifold}

\subsection{Matrix exponential parameterization of Special Orthogonal group}

\section{Unit sphere}

The unit sphere $\mathbb{S}^{N-1} \subset \mathbb{R}^N$ consists of all unit vectors $x$ of length 1, i.e. $x^\top x = 1$.
It is a special case of the Stiefel manifold.

Densities on the sphere are generally defined with respect to the spherically uniform measure.
This measure can be defined in many ways that are equivalent to within a constant factor that is not relevant for MCMC.
We will take the same approach as with the Stiefel manifold and define the spherically uniform measure as the measure that is invariant to transformations by $N \times N$ orthogonal matrices (rotations and reflections).
As discussed in the previous section, if $\tilde{X} \in O(N)$ is an orthogonal matrix whose first column is $x$, then $\tilde{X}^\top x = e_1$, where $e_1$ is the unit vector whose first element is 1, and $(\tilde{X}^\top \dv{x})_1 = 0$.
For convenience, we will assume that the rules used to produce $\tilde{X}$ enforce that $\det(\tilde{X}) = +1$.
The spherically uniform measure is then written
\[
    (\dv{x})^\wedge \equiv (\tilde{X}^\top \dv{x})^\wedge = \bigwedge_{i=2}^N (\tilde{X}^\top \dv{x})_i.
\]

\subsection{Embedding in real space using parameter expansion}

Let $f\colon y \mapsto (r, x)$ for $y \in \mathbb{R}^N$, $r > 0$, and $x \in \mathbb{S}^{N-1}$.
This map is bijective everywhere except $y=0$.

To compute the Jacobian determinant of the map, we first compute the differential form of $y=rx$:

\[\dv{y} = \dv{r} x + r \dv{x}.\]

Multiplying both sides by $\tilde{X}^\top$, we find

\[
    \tilde{X}^\top \dv{y} = \dv{r} \tilde{X}^\top x + r \tilde{X}^\top \dv{x} = \dv{r} e_1 + r (\tilde{X}^\top \dv{x}).
    \]

Then

\[
    (\tilde{X}^\top \dv{y})_i = \begin{cases}
        \dv{r} & \text{ if } i = 1 \\
        r (\tilde{X}^\top \dv{x})_i & \text{ if } i \ne 1
    \end{cases}.
\]

Note that $(\tilde{X}^\top \dv{y})^\wedge = \det({\tilde{X}}) (\dv{y})^\wedge = (\dv{y})^\wedge$.

Taking the exterior product of both sides, we then find

\[
\begin{aligned}
    (\dv{y})^\wedge &= \dv{r} \wedge \left(\bigwedge_{i=2}^N r (\tilde{X}^\top \dv{x})_i \right) = r^{N-1} \left(\dv{r} \wedge (\dv{x})^\wedge\right)\\
    \dv{r} \wedge (\dv{x})^\wedge &= r^{1-N} (\dv{y})^\wedge = \lVert y \rVert^{1-N} (\dv{y})^\wedge
\end{aligned}
\]

So the absolute Jacobian determinant is $\absdet{J_f(y)} = \lVert y \rVert^{1-N}$.

In the constrained space there is a free radial parameter $r$ that is unconstrained on the positive reals.
As a result, nothing prevents $r$ from growing arbitrarily large, which would cause the posterior to be improper, i.e. to have infinite mass.
Thus, we must also specify a prior $\pi(r)$.
While we may want to customize $\pi(r)$, a reasonable default choice is $r \sim \chi_N$, which has the density $\pi(r) \propto r^{N-1} \exp\left(-\frac{1}{2}r^2\right) $.

The combined log density correction is then
\[
  \absdet{J_f(y)} \pi(r) = \absdet{J_f(y)} \pi(\lVert y \rVert) \propto \lVert y \rVert^{1-N} \lVert y \rVert^{N-1} \exp\left(-\frac{1}{2}\lVert y \rVert^2\right) = \exp\left(-\frac{1}{2}\lVert y \rVert^2\right).
\]

That is, if $x$ is uniformly distributed on the sphere and $r$ follows this prior, then $y$ has a zero-centered standard multivariate normal distribution.

Since HMC is especially efficient at sampling normal distributions, this parameterization is very efficient when the marginal posterior distribution of $x$ is very diffuse on the sphere.
However, as the target distribution becomes concentrated on $x$, the geometry near the singularity $y=0$ may become problematic.

\subsection{Parameterization using hyperspherical coordinates}

Given $z \in \mathbb{R}^N$, $r > 0$ and angles $\phi=\{\phi_1, \ldots, \phi_{N-1}\}$, where $\phi_{N-1} \in (0, 2\pi]$ and $\phi_i \in (0, \pi]$ for $i < N-1$, we define the relationship between $z$ and $(r, \phi)$ as

\[
  z_i = \left(r \prod_{j=1}^{i-1} \sin(\phi_j)\right) \begin{cases}
    1 & \text{ if } i = N\\
    \cos(\phi_i) & \text{otherwise}
  \end{cases}.
\]

For the plane $N=2$ this works out the usual polar coordinate system:
\[
    \begin{aligned}
        z_1 &= r \cos\phi_1\\
        z_2 &= r \sin\phi_1
    \end{aligned}
\]

For $N=3$ this is \footnote{
    Note that this definition of hyperspherical coordinates differs for $N=3$ from the usual definition of spherical coordinates.
    It can be made the same through the coordinate permutation $(z_1, z_2, z_3) \to (z_2, z_3, z_1)$.
}
\[
    \begin{aligned}
        z_1 &= r \cos\phi_1\\
        z_2 &= r \sin\phi_1\cos\phi_2\\
        z_3 &= r \sin\phi_1\sin\phi_2
    \end{aligned}
\]

Our strategy is to compute the Jacobian determinant of the bijective map $g: (r, \phi) \mapsto z$ and relate it to the Jacobian determinant from the previous section to find the Jacobian determinant of the bijective map $f_2: \phi \mapsto x$ for $x \in \mathbb{S}^{N-1}$.

To do so, note first that
\[
  z_i^2 = \left(r^2 \prod_{j=1}^{i-1} \sin^2(\phi_j)\right) \begin{cases}
    1 & \text{ if } i = N\\
    \cos^2(\phi_i) & \text{otherwise}
  \end{cases}.
\]

Note also that
\[
    \begin{aligned}
        z_N^2 &= r^2 \prod_{j=1}^{N-1} \sin^2(\phi_j)\\
        z_{N-1}^2 + z_N^2 &= \left(r^2 \prod_{j=1}^{N-2} \sin^2(\phi_j)\right) \cos^2(\phi_{N-1}) + \left(r^2 \prod_{j=1}^{N-2} \sin^2(\phi_j) \right) \sin^2(\phi_{N-1}) = r^2 \prod_{j=1}^{N-2} \sin^2(\phi_j)\\
        z_{N-2}^2 + z_{N-1}^2 + z_N^2 &= \left(r^2 \prod_{j=1}^{N-3} \sin^2(\phi_j)\right) \cos^2(\phi_{N-2}) + \left(r^2 \prod_{j=1}^{N-3} \sin^2(\phi_j)\right) \sin^2(\phi_{N-2}) = r^2 \prod_{j=1}^{N-3} \sin^2(\phi_j)\\
        &\vdots \\
        \sum_{k=1}^i z_{N - i + k}^2 &= r^2 \prod_{j=1}^{N - i} \sin^2(\phi_j)\\
        &\vdots \\
        \sum_{k=1}^N z_{N - i + k}^2 &= r^2
    \end{aligned}
\]

We then differentiate both sides of this expression to get the differential form:

\[
    \begin{aligned}
        2 z_N \dv{z_N} &= 2r\dv{r} + 2 r^2 \left(\prod_{j=1}^{N-1} \sin^2(\phi_j)\right)\left( \sum_{j=1}^{N-1} \cot(\phi_j) \dv{\phi_j} \right)\\
        2z_{N-1}\dv{z_{N-1}} + 2z_{N}\dv{z_{N}} &= 2r \dv{r} + 2 r^2 \left(\prod_{j=1}^{N-2} \sin^2(\phi_j)\right) \left(\sum_{j=1}^{N-2} \cot(\phi_j)\right)\\
        2z_{N-2}\dv{z_{N-2}} + 2z_{N-1}\dv{z_{N-1}} + 2z_{N}\dv{z_{N}} &= 2r \dv{r} + 2 r^2 \left(\prod_{j=1}^{N-3} \sin^2(\phi_j)\right) \left(\sum_{j=1}^{N-3} \cot(\phi_j)\right)\\
        &\vdots \\
        2\sum_{k=1}^i z_{N - i + k} \dv{z_{N - i + k}} &= 2 r \dv{r} + 2 r^2 \left(\prod_{j=1}^{N - i} \sin^2(\phi_j)\right) \left(\sum_{k=1}^{N-i} \cot(\phi_k) \dv{\phi_k}\right)\\
        &\vdots \\
        2\sum_{k=1}^N z_{N - i + k} \dv{z_{N - i + k}} &= 2r\dv{r}
    \end{aligned}
\]

For each side separately, we now take the exterior product for all $i \in 1\ldots N$.
On the left side, we start at the top with $(2z_N dz_n) \wedge (2 z_{N-1} \dv{z_{N-1}} + 2 z_N \dv{z_N}) = 2^2 z_{N} z_{N-1} (\dv{z_N} \wedge \dv{z_{N-1}})$.
Note that the $\dv{z_N}$ from the second equation is eliminated since the exterior product already contains that term.
This trend continues as we proceed to the bottom, and the result is
\[
    \bigwedge_{i=1}^N 2z_{N-i+1} \dv{z_{N-i+1}} = \bigwedge_{i=1}^N 2z_i \dv{z_i} = 2^N \left(\prod_{i=1}^N z_i \right) (\dv{z})^\wedge
\]

Note that

\[
  \prod_{i=1}^N z_i = r \prod_{i=1}^{N-1} r \sin(\phi_i) \cos(\phi_i) \prod_{j=1}^{i-1} \sin(\phi_j) = r^N \prod_{i=1}^{N-1} \sin(\phi_i) \cos(\phi_i) \prod_{j=1}^{i-1} \sin(\phi_j)
\]

On the right side, we start at the bottom instead with
\[
(2r\dv{r} + 2 r^2 \sin(\phi_1) \cos(\phi_1) \dv{\phi_1}) \wedge (2r\dv{r}) = 4 r^3 \sin(\phi_1) \cos(\phi_1) (\dv{\phi_1} \wedge \dv{r}).
\]
Again, the $\dv{r}$ term in the second-from-bottom equation cancels since $\dv{r}$ is already included in the bottom term.
This trend also continues as we proceed to the top, and the result is then

\[
\begin{aligned}
    \left(\bigwedge_{i=1}^{N - 1} 2 r^2 \left(\prod_{j=1}^{N - i - 1} \sin^2(\phi_j)\right) \sin(\phi_{N-i}) \cos(\phi_{N-i}) \dv{\phi_{N-i}} \right) \wedge (2r\dv{r}) \\
    = 2^N r^{2N-1} \left(\prod_{i=1}^{N-1} \sin(\phi_{N-i}) \cos(\phi_{N-i}) \prod_{j=1}^{N-i-1} \sin^2(\phi_j)\right) \left((\dv{\phi})^\wedge \wedge \dv{r}\right)\\
    = 2^N r^{2N-1} \left(\prod_{i=1}^{N-1} \sin(\phi_i) \cos(\phi_i) \prod_{j=1}^{i-1} \sin^2(\phi_j)\right) \left((\dv{\phi})^\wedge \wedge \dv{r}\right)\\
    = 2^N r^{N-1} \left(\prod_{i=1}^N z_i\right) \left(\prod_{j=1}^{N-2} \sin(\phi_j)^{N-j-1}\right) \left((\dv{\phi})^\wedge \wedge \dv{r}\right)
\end{aligned}
\]

Equating the exterior products of each side, we have

\[
\begin{aligned}
    2^N \left(\prod_{i=1}^N z_i \right) (\dv{z})^\wedge &= 2^N r^{N-1} \left(\prod_{i=1}^N z_i\right) \left(\prod_{j=1}^{N-2} \sin(\phi_j)^{N-j-1}\right) \left((\dv{\phi})^\wedge \wedge \dv{r}\right)\\
    (\dv{z})^\wedge &= r^{N-1} \left(\prod_{j=1}^{N-2} \sin(\phi_j)^{N-j-1}\right) \left((\dv{\phi})^\wedge \wedge \dv{r}\right)
\end{aligned}
\]

As a result, the Jacobian determinant is

\[
    \absdet{J_g(r, \phi)} = r^{N-1} \prod_{i=1}^{N-2} \sin(\phi_i)^{N-i-1}
\]

As a quick check, for the circle ($N=1$), $\absdet{J_g(r, \phi)} = r$, and for the sphere ($N=2$), $\absdet{J_g(r, \phi)} = r^2 \sin(\phi_1)$, which are the expected results.
A similar result was given in \cite{muirhead2009aspects}.

Comparing with the previous section, we find

\[
\begin{aligned}
    (\dv{z})^\wedge &= r^{N-1} \left(\prod_{i=1}^{N-2} \sin(\phi_i)^{N-i-1}\right) \left((\dv{\phi})^\wedge \wedge \dv{r}\right) = r^{N-1} \left(\dv{r} \wedge (\dv{x})^\wedge \right) \\
    &= (-1)^{N-1} r^{N-1} \left((\dv{x})^\wedge \wedge \dv{r}\right)\\
    (\dv{x})^\wedge &= (-1)^{N-1}\left(\prod_{i=1}^{N-2} \sin(\phi_i)^{N-i-1}\right) (\dv{\phi})^\wedge
\end{aligned},
\]

so the absolute Jacobian determinant for the bijective map from hyperspherical coordinates to the sphere is
\[
    \prod_{i=1}^{N-2} \sin(\phi_i)^{N-i-1}.
\]

However, we're not done yet.
Recall that $\phi_j \in (0, \pi]$ for $i < N-1$ and $\phi_{N-1} \in (0, 2\pi]$.
So far we've essentially changed the problem of sampling on the hypersphere in $N$ dimensions to the problem of sampling on $N-1$ intervals, or, alternatively, $N-2$ hemicircles and 1 circle.

There are a few considerations when choosing the remaining parameterizations.
First, and most importantly, we want a smooth geometry.
Second, we want to avoid unnecessary work.
For example, if we choose parameter expansion for these parameters, we will end up with $N-1$ extra degrees of freedom.
Third, we want to avoid costly functions if possible.
For example, computing $\sin$ and $\cos$ is about 3 times more costly than computing $\exp$ and about 20 times more costly than computing a square.

\section{Cholesky factors of correlation matrices}

A $K \times K$ correlation matrix $\Omega$ is positive definite, has a unit diagonal, and all off-diagonal entries are between -1 and 1 (exclusive). Because it is positive definite, it can be Cholesky factored (see \ref{cholesky}) to a $K \times K$ lower-triangular matrix $L$
with positive diagonal elements such that $\Omega = L \cdot L^T$. Because the correlation matrix has a unit diagonal,

\[
  \Omega_{k,k} = L_k \cdot L_k^T = 1
\]

each row vector $L_k$ of the Cholesky factor is of unit length. The length and positivity
constraint allow the diagonal elements of L to be calculated from the off-diagonal
elements, so that a Cholesky factor for a $K \times K$ correlation matrix requires only $\binom{K}{2}$ unconstrained parameters.

\subsection{Hyperbolic tangent transform}

Suppose $y$ is a vector of $\binom{K}{2}$ unconstrained values. Let $z$ be a lower-triangular matrix with zero diagonal and below diagonal entries filled by row. For example, in the $3 \times 3$ case,
\[
z = \begin{bmatrix}
0 & 0 & 0\\
\tanh{y_1} & 0 & 0\\
\tanh{y_2} & \tanh{y_3} & 0\\
\end{bmatrix}
\]

The matrix $z$, with entries in the range $( -1, 1)$, is then transformed to the Cholesky factor $x$, by taking
\[
x_{i, j}=\left\{\begin{array}{lll}
0 & \text { if } i<j & \text { [above diagonal] } \\
\sqrt{1-\sum_{j^{\prime}<j} x_{i, j^{\prime}}^{2}} & \text { if } i=j & \text { [on diagonal] } \\
z_{i, j} \sqrt{1-\sum_{j^{\prime}<j} x_{i, j^{\prime}}^{2}} & \text { if } i>j & \text { [below diagonal] }
\end{array}\right.
\]

In the $3 \times 3$ case, this yields

\[
x=\left[\begin{array}{ccc}
1 & 0 & 0 \\
z_{2,1} & \sqrt{1-x_{2,1}^{2}} & 0 \\
z_{3,1} & z_{3,2} \sqrt{1-x_{3,1}^{2}} & \sqrt{1-\left(x_{3,1}^{2}+x_{3,2}^{2}\right)}
\end{array}\right]
\]

where the $z_{i,j}\in (-1, 1)$ are the tanh-transformed $y$.

\subsection{Absolute Jacobian determinant of inverse transform}

The Jacobian of the full transform is the product of the Jacobians of its component
transforms. First, for the inverse transform $z = \tanh{y}$, the derivative is
\[
\frac{d}{dy} \tanh y = \frac{1}{(\cosh y)^{2}}
\]
Second, for the inverse transform of $z$ to $x$, the resulting Jacobian matrix $J$ is of
dimension $\binom{K}{2} \times \binom{K}{2}$, with indexes $(i, j)$ for $(i > j)$. The Jacobian matrix is lower
triangular, so that its determinant is the product of its diagonal entries, of which
there is one for each $(i, j)$ pair.
\[
|\operatorname{det} J|=\prod_{i>j}\left|\frac{d}{d z_{i, j}} x_{i, j}\right|
\]

where

\[
\frac{d}{d z_{i, j}} x_{i, j}=\sqrt{1-\sum_{j^{\prime}<j} x_{i, j^{\prime}}^{2}}
\]

So the combined log Jacobian determinant
of the complete inverse transform is given by

\[
\log |\operatorname{det} J|=-2 \sum_{n \leq(\substack{K \\ 2}} \log \cosh y+\frac{1}{2} \sum_{i>j} \log \left(1-\sum_{j^{\prime}<j} x_{i, j^{\prime}}^{2}\right)
\]

\section{Positive definite matrices}

[SA: all transformations below can be generalized to SPD matrices of fixed rank, but we'd need to know what the common reference measure is for densities of SPD matrices, and I can't recall seeing one.]

An $N \times N$ positive definite (PD) matrix $X$ is a symmetric matrix that satisfies $v^\top X v > 0$ for all non-zero vectors $v$.
Covariance matrices are examples of PD matrices.

Density functions for distributions of PD matrices, such as the Wishart distribution, are generally defined with respect to the Lebesgue measure of the unique triangular elements $(\dv{X})^\wedge$.

\subsection{Cholesky parameterization}\label{cholesky}

Given a positive (semi-)definite matrix $X$ of rank $K$, we can uniquely decompose it as $X = L L^\top$ where $L$ is an $N \times N$ lower triangular matrix with $L_{jj} > 0$ for all $j \le K$ and $L_{ij} = 0$ for all $j > K$.
$L$ has $\frac{K(K+1)}{2} + (N - K)K$ degrees of freedom.
When $K=N$, $X$ is positive-definite.

We can construct $X$ by composing 2 bijective maps.
The first, $f_1$, enforces the constraint of the positive diagonal:

\[L_{ij} = f_1(Y)_{ij} = \begin{cases} \exp(Y_{ij}) & \text{ if } i = j \le K \\ Y_{ij} & \text{otherwise} \end{cases}\]

The Jacobian determinant for this map is $\absdet{J_{f_1}(Y)} = \exp(\sum_{i=1}^K Y_{ii})$.

The second map is $f_2: L \mapsto L L^\top = X$.
To compute the Jacobian of the map, we use the half-vectorization $\vectu(X)$.
Note that $X_{ij} = \sum_{k=1}^{\min(i, j)} L_{ik} L_{jk}$.
$X_{ij}$ is thus only dependent on elements of $L$ that are below and to the left of the $(i, j)$ position.
As a result, the Jacobian of $\vectu \circ f_2 \circ \vectu^{-1}$ is triangular, and its determinant is only the product of its diagonal elements: $\absdet{J_{f_2}} = \prod_{i=1}^N \prod_{j=1}^i \frac{\partial X_{ij}}{\partial L_{ij}}$.

In this index range, $\frac{\partial X_{ij}}{\partial L_{ij}} = (1 + \delta_{ij}) L_{ii}$.
As a result

\[
  \absdet{J_{f_2}(Z)} = 2^N \prod_{i = 1}^N \prod_{j = 1}^i L_{ii} = 2^N \prod_{i=1}^N |L_{ii}|^{N - i + 1} \\
\]

Composing the two bijective maps as $f = f_2 \circ f_1$, the resulting Jacobian determinant is

\[
  \absdet{J_{f}(Y)} = 2^N \exp\left( \sum_{i=1}^N (N - i + 2) Y_{ii}\right).
\]

\subsection{Matrix exponential parameterization}

For every PD matrix $X \in \mathbb{R}^{N \times N}$, there is a unique symmetric matrix $Y$ such that $X = \Exp(Y)$, where $\operatorname{Exp}(\cdot)$ is the matrix exponential, defined as
\[\Exp(Y) = \sum_{k=0}^\infty \frac{1}{k!} Y^k.\]

Let $Y = U \Lambda U^\top$ be the symmetric eigendecomposition of $Y$, where the matrix of eigenvectors $U$ is orthogonal, and the $\Lambda = \operatorname{diag}(\lambda)$ for $\lambda \in \mathbb{R}^N$.
Then $Y^2 = U \Lambda U^\top U \Lambda U^\top = U \Lambda^2 U^\top$.
In fact, $Y^c = U \Lambda^c U^\top$ for all real $c$.
As a result,
\[\Exp(Y) = U \Exp(\Lambda) U^\top,\]
where $\Exp(\Lambda)_{ij} = \exp(\lambda_i) \delta_{ij}$.

$\Exp(Y)$ is one example of a matrix function $f$, which we can roughly define for symmetric matrices as any function that obeys
\[f(Y) = U f(\Lambda) U^\top,\]
where $f(\Lambda)_{ij} = f(\lambda_i)\delta_{ij}$.
\footnote{
  In a similar way, we can generalize other scalar functions like the inverse, trigonometric, and hyperbolic functions to functions of matrices.
  All results in this section apply to these functions as well.
}

For such functions $f$, we can compute a common form of the Jacobian determinant.
Differentiating the expression, we find
\[
\begin{aligned}
  \dv{(f(Y))} &= \dv{U} f(\Lambda) U^\top + U f'(\Lambda) \dv{\Lambda} U^\top + U f(\Lambda) \dv{U}^\top\\
  U^\top \dv{(f(Y))} U &= U^\top \dv{U} f(\Lambda) + f(\Lambda) \dv{U}^\top U + f'(\Lambda)\dv{\Lambda},
\end{aligned}
\]
where $f'(\Lambda)_{ij} = \frac{\dv{(f(\lambda_i))}}{\dv{\lambda_i}}$.

Recall that $U^\top \dv{U}$ is skew-symmetric.
Note also that for skew-symmetrix $Z = -Z^\top$ and diagonal matrix $\operatorname{diag}(v)$,
\[(Z\operatorname{diag}(v) + \operatorname{diag}(v)Z^\top)_{ij} = Z_{ij} v_j + v_i Z_{ji} = Z_{ij} v_j - v_i Z_{ij} = Z_{ij} (v_j - v_i) = Z \circ V,\]
where $V_{ij} = v_j - v_i$ and $\circ$ is the Hadamard (elementwise) product.

From the right-hand side, note then that
\[(U^\top \dv{(f(Y))} U)_{ij} = \begin{cases} (U^\top \dv{U})_{ij} (f(\lambda_j) - f(\lambda_i)) & \text{ if } i \ne j \\ f'(\lambda_i) \dv{f(\lambda_i)} &\text{ if } i = j \end{cases}\]

The exterior product of the diagonal part is
\[
  \bigwedge_{i=1}^N f'(\lambda_i) \dv{\lambda_i} = \left(\prod_{i=1}^N f'(\lambda_i)\right) (\dv{\Lambda})^\wedge,
\]
while the exterior product of the sub-diagonal part is
\[
  \bigwedge_{i=1}^N \bigwedge_{j=i+1}^N (U^\top \dv{U})_{ij} (f(\lambda_j) - f(\lambda_i)) = \left( \prod_{i=1}^N \prod_{j=1}^{i-1} (f(\lambda_j) - f(\lambda_i)) \right) \left(U^\top \dv{U} \right)^\wedge
\]

Lastly, note for any symmetric $Z$ and orthogonal matrix $U$,
\[(U^\top \dv{Z} U)^\wedge = \det(U^\top)^N \det(U)^N (\dv{Z})^\wedge = (\dv{Z})^\wedge.\]

As a result, we can compute $(\dv(f(Y)))^\wedge$ by taking the exterior product of the sub-diagonal and diagonal parts: 
\[(\dv{(f(Y))})^\wedge = \left(\prod_{i=1}^N f'(\lambda_i) \prod_{j=1}^{i-1} (f(\lambda_j) - f(\lambda_i))\right) \left((U^\top \dv{U})^\wedge \wedge \dv{\Lambda}\right)\]

If $f$ is the identity map $f(Y) = Y$, then
\[ (\dv{Y})^\wedge = \left(\prod_{i=1}^N \prod_{j=1}^{i-1} (\lambda_j - \lambda_i)\right) \left((U^\top \dv{U})^\wedge \wedge \dv{\Lambda}\right)\]

Combining the latter two expressions:
\[(\dv{(f(Y))})^\wedge = \left(\prod_{i=1}^N f'(\lambda_i) \prod_{j=1}^{i-1} \frac{f(\lambda_j) - f(\lambda_i)}{\lambda_j - \lambda_i}\right) (\dv{Y})^\wedge\]

[REF, Edelman's notes] previously found this result.

For $X = f(Y) = \Exp(Y)$ then,
\[ (\dv{X})^\wedge = \left(\prod_{i=1}^N \exp(\lambda_i) \prod_{j=1}^{i-1} \frac{\exp(\lambda_j) - \exp(\lambda_i)}{\lambda_j - \lambda_i}\right) (\dv{Y})^\wedge,\]

so the Jacobian determinant of the exponential map is $\absdet{J_{\Exp}(Y)} = \prod_{i=1}^N \exp(\lambda_i) \prod_{j=1}^{i-1} \left|\frac{\exp(\lambda_j) - \exp(\lambda_i)}{\lambda_j - \lambda_i}\right|$.

Note that this expression is undefined when any eigenvalues are repeated.
This is not a problem for MCMC, since the eigenvalues are non-unique with probability 0, but non-unique eigenvalues could occur when maximizing a log-density.

Because $\lim_{h \to 0} \frac{f(x) - f(x + h)}{h} = f'(x)$, we can then modify the expression to handle such cases:
\[
  \absdet{J_{\Exp}(Y)} = \prod_{i=1}^N \prod_{j=1}^i \begin{cases}
    \left|\frac{\exp(\lambda_j) - \exp(\lambda_i)}{\lambda_j - \lambda_i}\right| &\text{ if } \lambda_i \ne \lambda_j \\ 
    \exp(\lambda_i) &\text{ if } \lambda_i = \lambda_j
  \end{cases}.
\]

If the eigenvalues are sorted in ascending order, then the log Jacobian determinant can be more stably computed as

\[
  \log\absdet{J_{\Exp}(Y)} = \sum_{j=1}^N j\lambda_j + \sum_{i=1}^j \begin{cases}
    \log(1 - \exp(\lambda_i - \lambda_j)) - \log(\lambda_j - \lambda_i) &\text{ if } \lambda_i \ne \lambda_j \\ 
    0 &\text{ if } \lambda_i = \lambda_j
  \end{cases}.
\]

>>>>

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{cite}
\usepackage[letterpaper, total={6.5in, 9in}]{geometry}
\usepackage{mathpazo}
\usepackage{sourcecodepro}
\usepackage{graphicx}

@@

@@

\subsection{Changes of Variables}

\subsection{Computing Jacobian}

@@

@@

@@

\section{Results}

\begin{figure}
    \centering
    \includegraphics[width=1.2\textwidth]{figures/ess.png}
    \caption{Effective Sample Size vs Total LeapFrog Steps}
    \label{fig:hyg}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=1.2\textwidth]{figures/rmse.png}
    \caption{Root Mean Squared Error vs Cumulative Leapfrog Steps}
    \label{fig:my_label}
\end{figure}
